Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-06-03 SIMPaper/SIMPaper.tex
finalized version for arxiv

\abstract{In survival analysis, covariate interactions are difficult to
model without prior knowledge using regression-based methods. Neural
network-based survival models have been developed for data-driven
interaction modeling, however, these approaches lack implementations for
time-varying interactions or a flexible baseline hazard. To address
this, we extend the case-base sampling framework and propose Case-Base
Neural Networks (CBNN), which introduce time directly into the model,
using commonly available neural network components for a simple
implementation. First, we case-base sample the survival data, converting
the survival time of each participant into discrete moments in time.
This allows us to estimate the probability an event occurs at a given
moment using a neural network to get the hazard. We compare CBNN to
survival methods based on regression and neural networks in two
simulations and two real data applications with a fixed neural network
architecture. We report metrics for each model over follow-up time using
both a proper (Index of Prediction Accuracy) and non-proper (inverse
probability censoring weights-adjusted concordance index) scoring rule.
A simple simulation shows CBNN outperforming the other neural network
methods. The complex simulation highlights the ability of CBNN to model
both a complex baseline hazard and time-varying interactions,
outperforming all competitors. The first real data application shows
CBNN outperforming all neural network competitors, while a second real
data application shows competitive performance. CBNN has shown that it
is more consistent in prediction performance across time in the
simulations and real data applications compared to the competing neural
network approaches.}

@@

The widespread use of the Cox proportional hazards model has led to a
focus on survival models based on hazard ratios and relative risks
rather than on survival curves and absolute risks \citep{hanley2009}. In
clinical settings, Cox models are frequently prioritized over
smooth-in-time accelerated failure time (AFT) models \citep{hanley2009},
which can estimate absolute risks by modeling the hazard directly
through user specification of a baseline hazard distribution. The user
may forgo this choice by implementing a flexible set of parameters using
splines, as proposed by Royston and Palmar \citep{royston2002flexible}.
These regression models also require user-specified interactions, which
may limit risk prediction accuracy.

Over longer follow-up, it is challenging to maintain the proportional
hazards assumption as the risk associated with each covariate may not be
constant with time. Previous studies related to breast cancer have
discovered time-varying interactions with covariates of interest, like
tumor size \citep{coradini2000time} \citep{hilsenbeck1998time}. Another
limitation of regression-based models is the requirement of prior
knowledge; the analyst must know potential time-varying interactions and
explicitly evaluate them.

Data-driven neural network approaches bypass the need for prior
knowledge of interaction terms. DeepSurv is a semi-parametric
proportional hazards neural network model that implements the Cox
partial log-likelihood as a custom loss function
\citep{katzman2018DeepSurv}. This results in a stepwise absolute risk
curve that cannot accommodate time-varying interactions. Compared to Cox
regression, DeepSurv has shown a small improvement in performance using
a single layer neural network on the Study to Understand Prognoses
Preferences and Risk Treatments (SUPPORT) dataset
\citep{knaus1995SUPPORT}. Farraggi et al.~suggested a modification to
the loss function required for a Cox neural network that can model
non-proportional hazards \citep{faraggi1995neural}. Wang et al.~took the
concept of extreme learning machines and applied them in a survival
context, removing the need for tuning the number of layers and nodes
\citep{wang2019extreme}. A Cox-based neural network combined with Local
Interpretable Model-Agnostic Explanations (LIME)
\citep{ribeiro2016model} interprets the association of each Single
Nucleotide Polymorphism (SNP) \citep{sun2020genome} to a phenotype.

DeepHit requires each survival time of interest to be specified in the
model and directly estimates survival curves, rather than deriving a
hazard function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and its concordance index (c-index)
score outperforms DeepSurv \citep{katzman2018DeepSurv} on the Molecular
Taxonomy of Breast Cancer International Consortium (METABRIC) dataset
\citep{curtis2012genomic}.

Ripley et al.~presented seven different distributions and modeling
paradigms based on neural networks \citep{ripley2004non}. They provided
the methodology behind parametric log-logistic and log-normal neural
network models. Like DeepHit, these models are restricted to one of the
seven baseline hazard distributions, reducing the flexibility of the
average survival prediction.

Deep Survival Machines (DSM) provides a flexible, parametric survival
model using neural networks with a mixture of distributions as the
baseline hazard \citep{dsmPaper}. DSM allows the user to specify a
mixture of distributions for a flexible baseline hazard. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

These methods need custom loss and activation functions that are
difficult to implement in practice. We note the complexity of these
alternative neural network approaches; DeepSurv, DeepHit, and DSM all
require custom loss functions \citep{katzman2018DeepSurv}
\citep{lee2018DeepHit} \citep{dsmPaper}, while DSM requires user
implementations for distributions beyond Log-Normal or Weibull and a
two-phase learning process \citep{dsmPaper}. Regression-based approaches
require prior specification of all interaction terms, which makes it
challenging to model covariate effects that change over time. In this
article, we propose Case-Base Neural Networks (CBNN) as a method that
models time-varying interactions and a flexible baseline hazard using
commonly available neural network components. Our approach to modeling
the full hazard uses case-base sampling \citep{hanley2009}, which allows
probabilistic models to predict survival outcomes. Following the
case-base framework, we can use transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by
including splines of time as covariates, we can approximate the
Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}. An alternative approach to Royston-Parmar
is case-base with logistic regression (CBLR), which provides some
flexibility in modeling, but still requires user specification of
time-varying interactions and baseline hazard, while CBNN can model both
without user specification; only a sufficiently deep network is
required.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choice and software implementation. Then, in Section
\ref{sims}, we describe our metrics and compare the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic
regression on simulated data. Next, Section \ref{casestudies} describes
the real-data analysis, while in Section \ref{discussion}, we explore
the implications of our results and contextualize them within neural
network survival analysis in a single event setting.

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics, and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. A
package is available for use at {[}GITHUBLINK{]}. The entire code base
to generate the analyses in this paper is available at {[}GITHUB
LINK{]}.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By allowing a neural network
to approximate a higher-order polynomial of time, the baseline hazard
specification is now data-driven, where user-defined hyperparameters
such as regularization, number of nodes, and layers control the
flexibility of the model. We provide a detailed description of the
choices we made in the next sub-section.

@@

Neural networks provide flexibility when defining the architecture and
optimization parameters of our models. These hyperparameter choices can
affect the estimated parameters and were chosen during a set of initial
simulations to determine if CBNN can learn complex interactions in
practice. We set a max number of epochs to be \(2000\), batch size as
\(512\), learning rate as \(10^{-3}\), decay as \(10^{-7}\), patience as
\(10\) epochs, \(\{50,50,25,25\}\) nodes in each hidden layer with
\(50\%\) dropout at each layer, a validation split of \(20\%\), a
testing split of \(10%
\), a minimum delta loss on the validation set of \(10^{-7}\) over 10
epochs, and Adam \citep{gulli2017} as our optimizer. These choices may
permit the model to approximate higher-order interactions while
preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test). We select
the best weights after training based on validation loss
\citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper} and the DeepHit model
using \textbf{pycox} \citep{lee2018DeepHit}. We made the components of
CBNN using the \textbf{keras} \citep{keras} package and the
\textbf{casebase} package for the sampling step. The \textbf{simsurv}
package \citep{simsurv} is used to simulate our simulation studies,
while \textbf{flexsurv} \citep{flexsurv} is used to fit a flexible
baseline hazard using splines for our complex simulation. We use the
implementation of c-ipcw from the python package \textbf{sksurv}
\citep{sksurv}. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the Brier score and IPA metrics.
We changed the \textbf{riskRegression} package to be used with any user
supplied risk function \(F\). To ensure that both R and Python-based
models are running in unison on the same data through our simulations
and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

The c-ipcw is a non-proper, rank-based metric that does not depend on
the censoring times in the test data \citep{uno2011}. The c-ipcw is
given by

\begin{align} \label{eq:cidx}
c-ipcw(t) &= \frac{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t) I\left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right)}{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t)},
\end{align} where the follow-up period of interest is (0,\(t\)),
\(I(\cdot)\) is an indicator function, \(\widehat{F}(X_{i},t)\) is the
risk function estimated for everyone in the study at time \(t\) and
c-ipcw can compare the performance of different models, where a higher
score is better. Note that the c-ipcw may produce misleading
performance, as it ranks based on survival times, not event status
\citep{cindexfails2019}. This metric is considered an unbiased
population concordance measure because of the IPCW adjustment
\citep{uno2011}.

\srb{in figure 2 i changed $C_{IPCW}$ to c-ipcw, as capital C is used for censoring. in the plots, would it make sense to leave it as a small letter?}

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/pmnnProjects/paper/../pmnnPlayground/reviewedAnalysisPMNN/figures/megaPlot}

@@

\caption{Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row shows the c-ipcw for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals).}\label{fig:megaPlot}
\end{figure}

@@

We simulate data from a simple model that primarily depends on a
constant baseline hazard:
\[h(t \mid X_i) = \lambda\cdot e^{\beta_{{1}}z_{1}+\beta_{{2}}z_{2}+\beta_{{3}}z_{3}},\]
with minimal covariate effects given by
\(\beta_{{1}}=0.1, \beta_{{2}}=0.1, \beta_{{3}}=0.1\) and
\(\lambda=1.0\). Once we simulate survival times, we introduce 10\%
random censoring.

@@

Figure \ref{fig:megaPlot} A, B and Table \ref{tab:megaTable} A show the
results for the simple simulation. The regression-based methods (CBLR,
Cox, Optimal) outperform the neural network ones in the simple
simulation setting. Among the neural network approaches, CBNN
outperforms all other methods in terms of both IPA and c-ipcw (Figure
\ref{fig:megaPlot} A, B). Specifically, we see CBNN is consistent across
time with smaller confidence intervals compared to DeepHit, DeepSurv and
DSM.

In this simple setting, the regression models are much closer to the
Optimal model, while the neural network models perform worse than the
null model. The wide confidence bands in c-ipcw suggest the neural
network models may be over-parameterized.

@@

This simulation demonstrates performance with the presence of a complex
baseline hazard and a time-varying interaction. Originally used to
demonstrate the spline-based hazard model proposed by Royston and Parmar
\citep{royston2002flexible}, the breast cancer dataset provides a
complex hazard from which we simulate, available in the
\textbf{flexsurv} R package \citep{flexsurv}. To increase the complexity
of our data-generating mechanism for this simulation, we design the
model as follows: \begin{align}
h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot z_{2} \cdot t)+ \tau_{2} ( z_{1} \cdot z_{3})+ \tau_{3} (z_{2} \cdot z_{3}), \nonumber
\end{align}

\FloatBarrier
\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Networks (CBNN), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Deep Survival Machines (DSM).}
\label{tab:megaTable}

\begin{center}\includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/pmnnProjects/paper/../pmnnPlayground/reviewedAnalysisPMNN/figures/megaTable} \end{center}

@@

Figure \ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B show the
performance over time on a test set in the complex simulation. Apart
from the Optimal regression model, CBNN outperforms the competing models
when examining IPA, and up to the 75-percentile of follow-up time for
c-ipcw. We expected the Optimal model to perform best in both metrics.
However, this was not the case for c-ipcw and may be due to an artifact
of concordance-based metrics, where a misspecified model may perform
better than a correctly specified one \citep{cindexfails2019}. We
attribute this difference in performance to the flexibility of CBNN in
terms of time-varying interactions and baseline hazard, flexibility the
other neural network models do not have, explaining their drop in
performance.

@@

Our complex simulation demonstrates the superior performance of CBNN in
ideal conditions with clean data. To obtain a more realistic performance
assessment, we compared models using two real datasets with a
time-to-event outcome. The first case study examines the SUPPORT dataset
\citep{knaus1995SUPPORT}. The second case study examines the METABRIC
dataset \citep{curtis2012genomic}. We use the same hyperparameters as in
the simulation studies. As we do not know the true model for the real
data, we exclude the Optimal model. We split the datasets keeping 20\%
of the observations as a test set. 20\% of the training set is kept
aside for validation at each epoch. We predict risk functions for
everyone in the test set, which are used to calculate our metrics. We
conduct 100 bootstrap re-samples for the real data applications to
obtain confidence intervals.

@@

The SUPPORT dataset tracks the time until death for seriously ill
patients at five American hospitals \citep{knaus1995SUPPORT}. We use a
pre-processed version of the dataset made available in the DeepSurv
package \citep{katzman2018DeepSurv}. This dataset contains 9104 samples
and 14 covariates (age, sex, race, number of comorbidities, presence of
diabetes, presence of dementia, presence of cancer, mean arterial blood
pressure, heart rate, respiration rate, temperature, white blood cell
count, serum's sodium, and serum's creatinine)
\citep{katzman2018DeepSurv}. Patients with missing features were
excluded and 68.10\% of the patients died during the 5.56-year study
period \citep{katzman2018DeepSurv}.

Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C
demonstrates the performance over time on a test set. The regression
models (CBLR, Cox) perform best considering IPA, followed by CBNN from
the \(25^{th}\) to \(100^{th}\) percentile of follow-up time. We note a
drop in performance for CBNN before the \(25^{th}\) percentile of
follow-up. For c-ipcw, CBNN outperforms the competing models
consistently over follow-up time. Note that performance is similar for
all models, aside from DeepSurv whose c-ipcw is lower than the rest
(Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C).

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers
of breast tumors, following 2000 individuals with breast cancer until
death \citep{curtis2012genomic}. They described these growths as primary
invasive breast carcinomas, with the goal of discovering both genetic
and clinical risk factors for breast cancer survival
\citep{curtis2012genomic}. We used the processed dataset made available
through DeepSurv \citep{katzman2018DeepSurv}, which includes 1980
patients of which 57.72\% die due to breast cancer within a median 10
years of follow-up \citep{katzman2018DeepSurv}. There are 11 covariates
in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR, and ERBB2) and 5
clinical features (hormone treatment indicator, radiotherapy indicator,
chemotherapy indicator, ER-positive indicator, and age at diagnosis)
\citep{katzman2018DeepSurv}.

>>>>

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While these methods have led
to better predictive performance than regression-based approaches, they
cannot model both time-varying interactions and complex baseline
hazards. To address this, we propose Case-Base Neural Networks (CBNN) as
a new approach that combines the case-base sampling framework with
flexible architectures. Our method naturally accounts for censoring,
does not require additional hyperparameters and uses binary
cross-entropy as the loss function. Using a novel sampling scheme and
data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. This study highlights the benefit of
combining case-base sampling with deep learning to provide a simple and
flexible modeling framework for data-driven, time-varying interaction
modeling of survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through a user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors, or a Cox model where the disease pathogenesis
may change with age, making it difficult to maintain the proportional
hazards assumption. For example, previous studies of breast cancer
incidence have discovered time-varying interactions with covariates of
interest, such as tumor size \citep{coradini2000time}. One approach to
provide flexibility in the baseline hazard involves using the basis of
splines on time in our model \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on the Study to Understand Prognoses Preferences and
Risk Treatments (SUPPORT) dataset \citep{knaus1995SUPPORT}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}. To remove the need for tuning the
number of layers and nodes, the concept of extreme learning machines has
been applied to a Cox neural network model \citep{wang2019extreme}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and it outperforms DeepSurv on the
Molecular Taxonomy of Breast Cancer International Consortium (METABRIC)
dataset \citep{curtis2012genomic}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a hyperparameter weighing its two
loss functions (negative log-likelihood and ranking losses) while DSM
requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without extra tuning
parameters.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choices and software implementation. Section
\ref{sims}, describes our metrics and compares the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic
regression (CBLR) on simulated data. Section \ref{casestudies} describes
the real-data analysis, while Section \ref{discussion} explores the
implications of our results and contextualizes them within neural
network survival analysis in a single event setting.

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
reproduce the figures and empirical results in this paper is available
at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the fukk hazard function. We provide a
detailed description of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a minimum delta loss on the validation set of \(10^{-7}\)
over 10 epochs and Adam \citep{gulli2017} as our optimizer. These
choices may permit the model to approximate higher-order interactions
while preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\) from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

The \(C_{IPCW}\) is a non-proper, rank-based metric that does not depend
on the censoring times in the test data \citep{uno2011}. The
\(C_{IPCW}\) is given by

\begin{align} \label{eq:cidx}
C_{IPCW}(t) &= \frac{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t) I\left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right)}{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t)}.
\end{align} where the follow-up period of interest is (0,\(t\)),
\(I(\cdot)\) is an indicator function, \(\widehat{F}(X_{i},t)\) is the
risk function estimated for everyone in the study at time \(t\) and
\(C_{IPCW}\) can compare the performance of different models, where a
higher score is better. Note that the \(C_{IPCW}\) may produce
misleading performance, as it ranks based on survival times, not event
status \citep{cindexfails2019}. This metric is considered an unbiased
population concordance measure because of the IPCW adjustment
\citep{uno2011}.

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/analyses/figures/megaPlot}

@@

\caption{Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals).}\label{fig:megaPlot}
\end{figure}

@@

We simulate data from a simple model that primarily depends on a
constant baseline hazard:
\[\log h(t \mid X_i) = \beta_{{1}}z_{1}+\beta_{{2}}z_{2}+\beta_{{3}}z_{3},\]
The covariate effects are given by
\(\beta_{{1}}=0.1, \beta_{{2}}=0.1, \beta_{{3}}=0.1\). Once we simulate
survival times, we introduce 10\% random censoring.

@@

Figure \ref{fig:megaPlot} A, B and Table \ref{tab:megaTable} A show the
results for the simple simulation. The regression-based methods (CBLR,
Cox, Optimal) outperform the neural network ones in the simple
simulation setting. Among the neural network approaches, CBNN
outperforms all other methods in terms of both IPA and \(C_{IPCW}\)
(Figure \ref{fig:megaPlot} A, B). Specifically, we see CBNN is
consistent across time with smaller confidence intervals compared to
DeepHit, DeepSurv and DSM.

In this simple setting, the regression models are much closer to the
Optimal model, while the neural network models perform worse than the KM
null model. The wide confidence bands in \(C_{IPCW}\) suggest the neural
network models may be over-parameterized.

@@

This simulation demonstrates performance with the presence of a complex
baseline hazard and a time-varying interaction. Originally used to show
the spline-based hazard model proposed by Royston and Parmar
\citep{royston2002flexible}, the breast cancer dataset provides a
complex hazard from which we simulate, available in the
\textbf{flexsurv} R package \citep{flexsurv}. To increase the complexity
of our data-generating mechanism for this simulation, we design the
model as follows: \begin{align}
\log h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot z_{2} \cdot t)+ \tau_{2} ( z_{1} \cdot z_{3})+ \tau_{3} (z_{2} \cdot z_{3}), \nonumber
\end{align}

\FloatBarrier
\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Network (CBNN), Optimal and Deep Survival Machines (DSM).}
\label{tab:megaTable}

\begin{center}\includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/analyses/figures/megaTable} \end{center}

@@

Figure \ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B show the
performance over time on a test set in the complex simulation. Apart
from the Optimal regression model, CBNN outperforms the competing models
when examining IPA and up to the 75-percentile of follow-up time for
\(C_{IPCW}\). We expected the Optimal model to perform best in both
metrics. However, this was not the case for \(C_{IPCW}\) and may be due
to an artifact of concordance-based metrics, where a misspecified model
may perform better than a correctly specified one
\citep{cindexfails2019}. We attribute the performance of CBNN to its
flexibility in modeling time-varying interactions and baseline hazard,
flexibility the other neural network models do not have.

@@

Our complex simulation demonstrates the superior performance of CBNN in
ideal conditions with clean data. To obtain a more realistic performance
assessment, we compared models using two real datasets with a
time-to-event outcome. The first case study examines the SUPPORT dataset
\citep{knaus1995SUPPORT}. The second case study examines the METABRIC
dataset \citep{curtis2012genomic}. We use the same hyperparameters as in
the simulation studies. As we do not know the true model for the real
data, we exclude the Optimal model. We split the datasets keeping 20\%
of the observations as a test set. 20\% of the training set is kept
aside for validation at each epoch. We predict risk functions for
everyone in the test set, which is used to calculate our metrics. We
conduct 100 bootstrap re-samples for the real data applications to
obtain confidence intervals.

@@

The SUPPORT dataset tracks the time until death for seriously ill
patients at five American hospitals \citep{knaus1995SUPPORT}. We use a
pre-processed version of the dataset made available in the DeepSurv
package \citep{katzman2018DeepSurv}. This dataset contains 9104 samples
and 14 covariates (age, sex, race, number of comorbidities, presence of
diabetes, presence of dementia, presence of cancer, mean arterial blood
pressure, heart rate, respiration rate, temperature, white blood cell
count, serum's sodium and serum's creatinine)
\citep{katzman2018DeepSurv}. Patients with missing features were
excluded and 68.10\% of the patients died during the 5.56-year study
period \citep{katzman2018DeepSurv}.

Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C
demonstrates the performance over time on a test set. The regression
models (CBLR, Cox) perform best considering IPA, followed by CBNN from
the \(25^{th}\) to \(100^{th}\) percentile of follow-up time. We note a
drop in performance for CBNN before the \(25^{th}\) percentile of
follow-up. For \(C_{IPCW}\), CBNN outperforms the competing models
consistently over follow-up time. Note that performance is similar for
all models, aside from DeepSurv whose \(C_{IPCW}\) is lower than the
rest (Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C).

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers
of breast tumors, following 2000 individuals with breast cancer until
death \citep{curtis2012genomic}. They described these growths as primary
invasive breast carcinomas, with the goal of discovering both genetic
and clinical risk factors for breast cancer survival
\citep{curtis2012genomic}. We used the processed dataset made available
through DeepSurv \citep{katzman2018DeepSurv}, which includes 1980
patients of which 57.72\% die due to breast cancer within a median 10
years of follow-up \citep{katzman2018DeepSurv}. There are 11 covariates
in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2) and 5
clinical features (hormone treatment indicator, radiotherapy indicator,
chemotherapy indicator, ER-positive indicator and age at diagnosis)
\citep{katzman2018DeepSurv}.
